Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-08 thesis/_book/_main.tex
started adding chapter 3

\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Conceptual and methodological problems with bias detection and avoidance in natural language processing},
            pdfauthor={Alicja Dobrzeniecka},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

@@

>>>>

\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Conceptual and methodological problems with bias detection and avoidance in natural language processing},
            pdfauthor={Alicja Dobrzeniecka},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{todonotes}

@@

\chapter{Bayesian analysis of cosine-based
bias}\label{bayesian-analysis-of-cosine-based-bias}

We start with loading the libraries needed for the analysis.

\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(ggthemes)}
\KeywordTok{library}\NormalTok{(rethinking)}
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(ggpubr)}
\KeywordTok{library}\NormalTok{(kableExtra)}
\KeywordTok{library}\NormalTok{(dplyr)}
\KeywordTok{library}\NormalTok{(ggExtra)}
\KeywordTok{library}\NormalTok{(cowplot)}
\end{Highlighting}
\end{Shaded}

\normalsize

\chapter{Walkthrough with the religion
dataset}\label{walkthrough-with-the-religion-dataset}

We will use the choice of protected words and stereotypical predicates
used in REF. This is a decent point of departure, not only we want to
compare our method to that of REF, but also because this data format is
fairly general (as contrasted, say, with a set up for binary
stereotypes). Note also that the method we develop here can fairly
easily be run for different stereotypization patterns. Let's start with
explaining the method and its deployment using a dataset obtained for
the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we
prepared. In order to obtain this dataset, we calculated the cosine
distance between each protected word and each word from both the
bias-related attribute groups, which were used in the original study,
and to neutral and human control attributes which we added as control
groups. For instance, for religion, the bias-related predicates (coming
from the original study in REF) include muslim bias attributes, jew bias
attributes, christian bias attributes (see a list in the APPENDIX).

We decided to add control groups in the form of two classes --- neutral
words and human-related words. Without a proper control group it is
quite hard to compare the resulting cosine distances and decide on their
significance in bias detection. We prepared approximately
300\todo{FIX later} more or less neutral words to double-check the
prima-facie neutral hypothesis that their cosine similarity to the
protected words will oscillate around 0 (that is, the distances will be
around 1). This provides us with a more reliable point of reference.
Moreover, we added human attributes that are associated with people in
general to investigate whether the smaller cosine distance between
protected words and stereotypes can result simply from the fact that the
stereotype predicates are associated with humans. For two control
groups, we have randomly drawn 300 words that do not express any
property usually attributed to humans, and human attributes.
\todo{describe once the words are selected}

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:religionTableHead}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}

\normalsize

The \texttt{protectedWord} column contains words from a protected class
that (in a perfect world according to the assumptions of the orignal
study) should not be associated with harmful stereotypes.
\texttt{wordToCompare} contains attributes, including stereotypes and
control group words. For each row we compute the cosine distances
between a given protected word and a given attribute word.
\texttt{wordClass} tells us which class an attribute is supposed to be
stereotypically associated with, that is, whether the word from
\texttt{wordToCompare} is associated stereotypically with jews,
christians or muslims, or whether it belongs to a control group.
\texttt{cosineDistance} is simply a calculation of the cosine distance
between protected word and atrribute. \texttt{cosineSimilarity} contains
the result of substracting cosine distance from 1. \texttt{connection}
contains information about the relation type between a protected word
and an attribute. If the attribute is e.g.~a harmful jewish stereotype
and the protected word is also from the judaism group, the connection
has value \texttt{associated}. If the attribute is still stereotypically
jewish, but the protected word comes from another religion, the
connection is labelled as \texttt{different}. If the attribute belongs
to a neutral group then the connection is labelled as \texttt{none} and
if an attribute belongs to the \texttt{human} class, then the connection
is labelled as \texttt{human}.

First let's take a look at the empirical distribution of distances by
the connection type, initially ignoring the human control class for now.

\vspace{1mm} \footnotesize

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}

\normalsize

The first impression is that while there is a shift for associated words
towards smaller cosine distances as compared to the neutral words,
slightly surprisingly a slightly weaker shift in the same direction is
visible for attributes associated with different stereotypes. Moreover,
the empirical distributions overlap to alarge extent and the means
grouped by connection type do not seem too far from each other. In fact,
as there is a lot of variety in the consine distances (as we will soon
see), abd we need to gauge the uncertaintly involved, and to look more
carefully at individual protected words to get a better idea of how the
cosine distance distribution changes for different attribute groups and
different protected classes. Now, let's add the human attributes to the
picture:

\vspace{1mm} \footnotesize

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}

\normalsize

\noindent Notice that the distribution for \texttt{human} (even though
we did our best not to include in it any stereotype-related atributes)
is left-skewed, with much overlap with \texttt{associated} and
\texttt{different}, which illustrates the need to take being associated
with humans as an important predictor.

Our focus lies in \texttt{connection} as a predictor. Morever, later on
we'll be interested in looking at the protected words separately, and at
protected words split by connection. For technical reasons it is useful
to represent these factors as integer vectors.

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
\end{Highlighting}
\end{Shaded}

\normalsize

A short script, \texttt{cleanDataset} to make this faster, so
equivalently:

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\normalsize

For now, let's focus on five protected words related to islam (``imam'',
``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
associates with islam four stereotypical attributes (``violent'',
``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
the empirical distributions for these protected words.

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}

\normalsize

\noindent Once we focus on words related to islam, the associated bias
seems to be stronger than in the whole dataset. This is a step towards
illustrating that the distribution of bias is uneven.

Now, say we want to look at a single protected word. Since the dataset
also contains comparison multiple control neutral and human attributes,
we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
control groups of those for the visualisation purposes.

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}

\CommentTok{#we plug in our visualisation script}
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
\CommentTok{#two arguments: dataset and protected word}
\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}

\normalsize

Note that the distance between the grey point and the other points is
proportional to cosine distance, the non-grey point size is proportional
to cosine similarity to the protected word, and color groups by the
connection type. So for \texttt{muslim} it seems that the stereotypes
coming from the word list are fairly well visible. To give you some
taste of how uneven the dataset is, compare this to what happens with
\texttt{priest}.

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}

\CommentTok{#we plug in our visualisation script}
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
\CommentTok{#two arguments: dataset and protected word}
\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}

\normalsize

\noindent Here you can see that some human attributes are closer than
stereotype attributes, and that there is no clear reason to claim that
\texttt{associated} attributes are closer than \texttt{different} or
\texttt{human} attributes. This, again, illustrates the need of
case-by-case analysis with control groups.

The general idea now is that the word lists provided in different pieces
of research are just samples of attributes associates with various
stereotypes and should be treated as such: the uncertaintly involved and
the sample sizes should have clear impact on our estimates.

We will now think of cosine distance as the output variable, and will
build a few bayesian models to compare. First, we just build a baseline
model which estimates cosine distance to the attributes separately for
each protected word. The underlying idea is that different protected
words migh in general have different relations to all the attributes and
this relations should be our point of departure.

Here is the intution behind the mathematical Bayesian model involved.
Our outcome variable is \texttt{cosine\ difference}, which we take to me
normally distributed around the predicted mean for a given protected
word (that is, we assume the residuals are normally distributed). The
simplest model specification is:

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} \\
m_{pw} & ~ dnorm(1,.5) \\
\sigma &\sim  dcauchy(0,1)
\end{align}

That is, we assume the estimated means might be different for diferent
protected words and our prior for the mean and the overal standard
deviation are normal with mean 1 and sd=.5 and half-cauchy with
parameters \texttt{0,1}. Further on we'll also estimate additional
impact the connection type may have. For this impact we take a slightly
skeptical prior centered around 0 distributed normally with sd = 1.
These are fairly weak and slightly skeptical regularizing priors, which
can be illustrated as follows:

\vspace{2mm}

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}

Now we can define and compile the baseline model. Its parameters will
have a posterior distribution obtained using either Hamiltionian Monte
Carlo methods (STAN) available through the \texttt{rethinking} package.

### 2021-06-08 thesis/index.Rmd
started adding chapter 3

---
title: "Conceptual and methodological problems with bias detection and avoidance in natural language processing"
author: "Alicja Dobrzeniecka"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_book: 
    keep_tex: yes
    toc_depth: 6
    latex_engine: xelatex
  bookdown::gitbook:
    lib_dir: "book_assets"
includes:
    in_header:
      - Rafal_latex6.sty
documentclass: book
classoptions: dvipsnames, usenames
site: bookdown::bookdown_site
bibliography: [../references/bibliographyRafal.bib]
biblio-style: 
link-citations: yes
csl: apa-5th-edition.csl
---

>>>>

---
title: "Conceptual and methodological problems with bias detection and avoidance in natural language processing"
author: "Alicja Dobrzeniecka"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_book: 
    keep_tex: yes
    toc_depth: 6
    latex_engine: xelatex
    extra_dependencies: ["todonotes"]
  bookdown::gitbook:
    lib_dir: "book_assets"
includes:
    in_header:
      - Rafal_latex6.sty
documentclass: book
classoptions: dvipsnames, usenames
site: bookdown::bookdown_site
bibliography: [../references/bibliographyRafal.bib]
biblio-style: 
link-citations: yes
csl: apa-5th-edition.csl
---
